Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-12-20 paper-bedpca.tex
add future directions

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
We expect population structure to be well distributed along the genome so that we get loadings (weights for each variant for each PC) that are normally distributed around 0 (with small effect sizes).
In contrast, long-range LD structure is basically capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
X-axis represent positions of variants in the data, and y-axis the value of loadings.
Points are hex-binned to make plotting of such large number of points manageable.
\label{fig:UKBB-loadings40}}
\end{figure}

@@

We summarise our work into several recommendations for computing PCA, and propose the pipeline shown in figure \ref{fig:pipeline}. Note that we have not included standard steps such as initial quality control filters and post-analysis checks (e.g.\ visual inspection of different plots).
This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}. 
Note that one should remove both individuals in each pair of related individuals. This ensures that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
To make these steps more computationally efficient, we explored solutions for not recomputing PCA from scratch when removing a few samples or variants. Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}. 
Once PCA is done, one should look at the loadings to make sure that there is no visible LD structure left (peaks in PC loadings as shown in figures \ref{fig:UKBB-loadings40} and \ref{fig:UKBB-loadings}), since including PCs that capture LD as covariates in genetic analyses can lead to reduced power for variants in these regions. 
These long-range LD regions are captured by PCs because variants in these regions are extensively correlated (even after a first step of LD thinning), and PCA just captures the variance in these correlated variants. 
This is a similar problem as including related individuals or even whole families in PCA, they would show up as structure in PCs. 
One should also look at PC scores to identify which PCs capture population structure.
As in many applications, we believe there should be a compromise between signal and noise. Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that indicate LD (e.g.\ PC19-40 in figure \ref{fig:UKBB-loadings40}).

When analysing a dataset that is composed mainly of one population (e.g.\ British people in the UK Biobank), we found that it is useful to subset these individuals to reduce the imbalance between the different population sizes.
Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
This suggests that designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 different populations, are highly relevant for capturing population structure \cite[]{10002015global}.

@@

We tackled many problems related to computing PCA on genotype data in this paper. 
Yet, more work is still needed. 
For example, the effect of including long-range LD regions in GWAS should be investigated quantitatively.
These are directions of future work.

[TODO: ADD PART ON MORE WORK NEEDED -> optimal subsetting for populations and effect of including long-range LD regions in e.g.\ GWAS.]

@@

>>>>

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
X-axis represent positions of variants in the data, and y-axis the value of loadings.
Points are hex-binned to make plotting of such large number of points manageable.
\label{fig:UKBB-loadings40}}
\end{figure}

@@

We summarise our work into several recommendations for computing PCA, and propose the pipeline shown in figure \ref{fig:pipeline}. Note that we have not included standard steps such as initial quality control filters and post-analysis checks (e.g.\ visual inspection of different plots).
This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}. 
Note that one should remove both individuals in each pair of related individuals. This ensures that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
To make these steps more computationally efficient, we explored solutions for not recomputing PCA from scratch when removing a few samples or variants. Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.

Once PCA is done, one should check the PC scores (scores of each individual for each PCA dimension) and PC loadings (weights for each variant for each PCA dimension). 
We differentiate PCs in 3 broad types: the ones capturing LD structure, the ones capturing population structure, and noise.
We expect population structure to be well distributed along the genome so that we get loadings that are normally distributed around 0 (with small effect sizes).
In contrast, long-range LD structure is basically capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
Therefore, PCs capturing LD structure can be identified by looking for peaks in PC loadings (e.g.\ PC17-20 in figure \ref{fig:UKBB-loadings}). 
To identify which PCs capture population structure, and which ones are probably just noise, one should also look at PC scores (colored by ancestry if possible).
As in many applications, we believe a compromise between signal and noise should be preferred. Therefore, we recommend using only PCs that show structure (e.g.\ PC1-14, and probably also PC15-16, in figure \ref{fig:UKBB-scores}) and excluding PCs that does not seem to capture any population structure (e.g.\ PC17-20 in figure \ref{fig:UKBB-scores}).

When analysing a dataset that is composed mainly of one population (e.g.\ British people in the UK Biobank), we found that it is useful to subset these individuals to reduce the imbalance between the different population sizes.
Likewise, previous works have shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
Determining how much overdominant populations should be subsampled to maximise population structure captured by PCA is a direction of future work.
The remaining individuals can then be projected onto the resulting PCA space using the OADP projection we recommend in this paper.
This suggests that designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 different populations, are highly relevant for capturing population structure \cite[]{10002015global}.

@@

The ubiquitous use of PCA in a wide variety of genomic analyses makes it difficult to establish universal guidelines for such analysis. 
Although we have tackled many problems related to computing PCA on genotype data in this paper, we do not answer other important problems, such as how to best control for population structure in genomic analyses. 
For example, when conducting a GWAS, should one restrict to a homogenous sample, or is it enough to just include PCs that capture population structure as covariates, or should one also use PCs as covariates in mixed linear models \cite[]{price2010new,loh2015efficient}?
Similarly, in some analyses, it may be beneficial to include PCs that capture long-range or even inter-chromosomal LD. More work is needed to understand these fundamental problems, and to provide precise guidelines for conducting successful GWAS, heritability, and other genomic analyses where PCA is used. These are directions of future work.

@@

%\clearpage
\vspace*{5em}
